Flooding of the Okavango Delta influences Connectivity for Dispersing African Wild Dogs

### 2023-03-14 Chapter_1/04_Manuscript/Manuscript.tex
upd
various updates

\textsuperscript{1} Department of Evolutionary Biology and Environmental
Studies, University of Zurich, Winterthurerstarsse 190, 8057 Zurich,
Switzerland.

\textsuperscript{2} Botswana Predator Conservation Program, Private Bag 13,
Maun, Botswana.

@@

% Issues with Both Methods
% \subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be justifiable for
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory towards an unknown endpoint
\citep{Koen.2014, Cozzi.2020}. CT, on the contrary, posits that animals move
according to a random walk, entailing that autocorrelation between subsequent
movements cannot be rendered \citep{Diniz.2019}. For dispersers, however,
autocorrelated movements are regularly observed \citep{Cozzi.2020,
Hofmann.2021}, meaning that dispersal trajectories are usually strongly
directional. An interesting generalization that bridges the continuum between
LCPA and CT has been proposed by \citep{Panzacchi.2016} and enables to
capitalize on the merits of both approaches. Despite these and several other
generalizations of LCPA and CT (e.g. \citealp{Pinto.2009, Landguth.2012,
Panzacchi.2016, Brennan.2020}), some shortcomings remain. Most notably, all of
these methods rely on static permeability or resistance surfaces that can't
reflect the temporal dimension of dispersal. This permits statements about the
expected duration for moving between habitat patches \citep{Martensen.2017,
Diniz.2019}.

@@

To ensure reliable connectivity estimates, we determined the number of simulated
dispersal trajectories required to reach a ``steady state''. For this purpose,
we distributed 1,000 rectangular ``checkpoints'', each with an arbitrary extent
of 5 km x 5 km, at random coordinates within the study area (excluding the
buffer). We then determined the relative frequency at which each checkpoint was
traversed by simulated dispersal trajectories (hereafter referred to as relative
traversal frequency) as we gradually increased the number of simulated
trajectories from 1 to 50,000. To assess variability in the relative traversal
frequency, we repeatedly subsampled 100 times from all 50'000 trajectories and
computed the mean traversal frequency across replicates, as well as its 95\%
prediction-interval for each checkpoint. We considered connectivity to have
reached a steady state once the width of the prediction-interval dropped below a
value of 0.01 for all checkpoints.

@@

\subsubsection{Validation}
To validate our predictions of connectivity, we utilized additional dispersal
data that was collected on eight dispersing coalitions between 2019 and 2022
(totalling to 2'668 GPS locations). We used path selection analysis (PSF,
\citealp{Cushman.2010}) to assess if observed dispersal trajectories followed
areas of high predicted connectivity. Similar to SSF, PSF enables to detect
selection for certain features by comparing observed paths to randomly generated
paths. Here, we paired each observed path with 50 random paths that we generated
by randomly rotating and shifting observed paths by a random angle \(\alpha \sim
U(-\pi, +\pi)\) and a random distance \(d \sim U(0 \) km, \(50\) km\(\)). Along
each path, we then extracted connectivity values from the heatmap (see above)
generated after 68, 125, 250, 500, and 2'000 simulated steps, respectively.
Finally, we ran conditional logistic regression to contrast observed and random
paths. In case of systematic selection for high-connectivity areas, the
regression coefficients from the corresponding conditional logistic regression
model should be positive.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{99_MovementModel}
    \caption{(a) Most parsimonious movement model for dispersing wild dogs. The
    model comprises a habitat kernel, a movement kernel, as well as their
    interactions. The horizontal line segments delineate the 90\%, 95\%, and
    99\% confidence-intervals for the respective \(\beta\)-coefficients.
    Significance codes: * \(p < 0.10\), ** \(p < 0.05\), *** \(p < 0.01\). (b)
    Results from the k-fold cross validation procedure. Subfigure b1 shows rank
    frequencies of realized steps according to model predictions with known
    preferences, whereas subfigure b2 shows rank frequencies of realized steps
    when assuming random preferences. The blue ribbon shows the prediction
    interval around a loess smoothing regression that we fitted to ease the
    interpretation of the plots. The significant correlation between rank and
    associated frequency in (b1) highlights that the most parsimonious model
    successfully outperformed a random guess (b2) and frequently assigned low
    ranks (i.e. high selection scores) to realized steps but only rarely high
    ranks (i.e. low selection scores). (c) Results from the PSF analysis using
    independent dispersal data show that dispersers preferrably moved through
    areas where our heatmaps predicted high connectivity. Results are shown for
    heatmaps realized after 68, 125, 250, 500, and 2'000 simulated steps,
    respectively.}
    \label{MovementModel}
  \end{center}
\end{figure}

\subsection{Dispersal Simulation}
Dispersal simulations based on the most parsimonious movement model proved
useful for assessing landscape connectivity. Of the 50,000 simulated dispersal
trajectories that originated from the main study area, only 4.5\% reached a map
boundary, suggesting that we successfully mitigated biases from boundary
effects. Moreover, our examination of the relative traversal frequency across
all checkpoints showed that the relative traversal frequency reached a steady
state after 10,500 simulated dispersal trajectories (Figure S4). Although
variability in the relative traversal frequency kept decreasing as we increased
the number of simulated dispersers, the marginal benefit of simulating
additional trajectories diminished quickly (Figure S4).

@@

% \subsection{Disadvantages of ISSF Simulations}
Despite the many benefits and great flexibility offered by simulations from
ISSFs, one must also be aware of the associated limitations. For example, while
our approach of simulating dispersal proved usedful to assess landscape
connectivity, it was computationally costly. Simulating 80,000 dispersal
trajectories for 2'000 steps across the KAZA-TFCA required five days of
computation on a regular desktop machine (AMD Ryzen 7 2700X octa-core processor
with 3.6 GHz, 64 GB of RAM). The long simulation time was primarily caused by
the massive extent of the study area considered (ca. 1.3 Mio
km\textsuperscript{2}), the large number of simulated trajectories, and the fact
that we extracted covariates along each step, rather than just at their start or
endpoints. Most connectivity studies focus on smaller study areas (e.g.
\citealp{Kanagaraj.2013, Clark.2015, McClure.2016, Abrahms.2017, Zeller.2020})
and will therefore require fewer simulations and achieve faster simulation times
(given the same spatial resolution). We also believe that fewer simulated
trajectories will often suffice, as the relative traversal frequency by
simulated trajectories through randomly placed checkpoints across our study area
converged already after 10,500 runs. The exact number of required simulations to
achieve reliable estimates of connectivity will, of course, vary depending on
the structure of the landscape and the dispersal capabilities of the focal
species \citep{Gustafson.1996}. For species that disperse short distances
through homogeneous environments, few simulations may suffice to gauge
connectivity, whereas for species that disperse over long distances through
heterogeneous habitats, a large number of simulations will be required to
sufficiently explore the spectrum of possible routes. Finally, it may often
suffice to extract covariates at each step's start or endpoints, thus
considerably speeding up simulation times \citep{Signer.2017}.

>>>>

\textsuperscript{1} Department of Evolutionary Biology and Environmental
Studies, University of Zurich, Winterthurerstrasse 190, 8057 Zurich,
Switzerland.

\textsuperscript{2} Botswana Predator Conservation Program, Wild Entrust,
Private Bag 13, Maun, Botswana.

@@

% Issues with Both Methods
% \subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be justifiable for
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory towards an unknown endpoint
\citep{Koen.2014, Cozzi.2020}. CT, on the contrary, posits that animals move
according to a random walk, entailing that autocorrelation between subsequent
movements cannot be rendered \citep{Diniz.2019}. For dispersers, however,
autocorrelated movements are regularly observed \citep{Cozzi.2020,
Hofmann.2021}, meaning that dispersal trajectories are usually strongly
directional. An interesting generalization that bridges the continuum between
LCPA and CT has been proposed by \cite{Panzacchi.2016} and enables to
capitalize on the merits of both approaches. Despite these and several other
generalizations of LCPA and CT (e.g. \citealp{Pinto.2009, Landguth.2012,
Panzacchi.2016, Brennan.2020}), some shortcomings remain. Most notably, all of
these methods rely on static permeability or resistance surfaces that can't
reflect the temporal dimension of dispersal. This permits statements about the
expected duration for moving between habitat patches \citep{Martensen.2017,
Diniz.2019}.

@@

To ensure reliable connectivity estimates, we determined the number of simulated
dispersal trajectories required to reach a ``steady state''. For this purpose,
we distributed 1,000 rectangular ``checkpoints'', each with an arbitrary extent
of 5 km x 5 km, at random coordinates within the study area (excluding the
buffer). We then determined the relative frequency at which each checkpoint was
traversed by simulated dispersal trajectories (hereafter referred to as relative
traversal frequency) as we gradually increased the number of simulated
trajectories from 1 to 50'000. To assess variability in the relative traversal
frequency, we repeatedly subsampled 100 times from all 50'000 trajectories and
computed the mean traversal frequency across replicates, as well as its 95\%
prediction-interval for each checkpoint. We considered connectivity to have
reached a steady state once the width of the prediction-interval dropped below a
value of 0.01 for all checkpoints.

@@

\subsubsection{Validation}
To validate our predictions of connectivity, we utilized additional dispersal
data that was collected on eight dispersing coalitions between 2019 and 2022
(totalling to 2'668 GPS locations). We used a path selection function (PSF,
\citealp{Cushman.2010}) to assess if observed dispersal trajectories followed
areas of high predicted connectivity. Similar to SSF, PSF enables to detect
selection for certain features by comparing observed paths to randomly generated
paths. Here, we paired each observed path with 50 random paths that we generated
by randomly rotating and shifting observed paths by a random angle \(\alpha \sim
U(-\pi, +\pi)\) and a random distance \(d \sim U(0 \) km, \(50\) km\(\)). Along
each path, we then extracted connectivity values from the heatmap (see above)
generated after 68, 125, 250, 500, and 2'000 simulated steps, respectively.
Finally, we ran conditional logistic regression to contrast observed and random
paths. In case of systematic selection for high-connectivity areas, the
regression coefficients from the corresponding conditional logistic regression
model should be positive.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{99_MovementModel}
    \caption{(a) Most parsimonious movement model for dispersing wild dogs. The
    model comprises a habitat kernel, a movement kernel, as well as their
    interactions. The horizontal line segments delineate the 90\%, 95\%, and
    99\% confidence-intervals for the respective \(\beta\)-coefficients.
    Significance codes: * \(p < 0.10\), ** \(p < 0.05\), *** \(p < 0.01\). (b)
    Results from the k-fold cross-validation procedure. Subfigure b1 shows rank
    frequencies of realized steps according to model predictions with known
    preferences, whereas subfigure b2 shows rank frequencies of realized steps
    when assuming random preferences. The blue ribbon shows the prediction
    interval around a loess smoothing regression that we fitted to ease the
    interpretation of the plots. The significant correlation between rank and
    associated frequency in (b1) highlights that the most parsimonious model
    successfully outperformed a random guess (b2) and frequently assigned low
    ranks (i.e. high selection scores) to realized steps but only rarely high
    ranks (i.e. low selection scores). (c) Results from the PSF analysis using
    independent dispersal data show that dispersers preferrably moved through
    areas where our heatmaps predicted high connectivity. Results are shown for
    heatmaps realized after 68, 125, 250, 500, and 2'000 simulated steps,
    respectively.}
    \label{MovementModel}
  \end{center}
\end{figure}

\subsection{Dispersal Simulation}
Dispersal simulations based on the most parsimonious movement model proved
useful for assessing landscape connectivity. Of the 50'000 simulated dispersal
trajectories that originated from the main study area, only 4.5\% reached a map
boundary, suggesting that we successfully mitigated biases from boundary
effects. Moreover, our examination of the relative traversal frequency across
all checkpoints showed that the relative traversal frequency reached a steady
state after 10'500 simulated dispersal trajectories (Figure S4). Although
variability in the relative traversal frequency kept decreasing as we increased
the number of simulated dispersers, the marginal benefit of simulating
additional trajectories diminished quickly (Figure S4).

@@

% \subsection{Disadvantages of ISSF Simulations}
Despite the many benefits and great flexibility offered by simulations from
ISSFs, one must also be aware of the associated limitations. For example, while
our approach of simulating dispersal proved usedful to assess landscape
connectivity, it was computationally costly. Simulating 80'000 dispersal
trajectories for 2'000 steps across the KAZA-TFCA required five days of
computation on a regular desktop machine (AMD Ryzen 7 2700X octa-core processor
with 3.6 GHz, 64 GB of RAM). The long simulation time was primarily caused by
the massive extent of the study area considered (ca. 1.3 Mio
km\textsuperscript{2}), the large number of simulated trajectories, and the fact
that we extracted covariates along each step, rather than just at their start or
endpoints. Most connectivity studies focus on smaller study areas (e.g.
\citealp{Kanagaraj.2013, Clark.2015, McClure.2016, Abrahms.2017, Zeller.2020})
and will therefore require fewer simulations and achieve faster simulation times
(given the same spatial resolution). We also believe that fewer simulated
trajectories will often suffice, as the relative traversal frequency by
simulated trajectories through randomly placed checkpoints across our study area
converged already after 10'500 runs. The exact number of required simulations to
achieve reliable estimates of connectivity will, of course, vary depending on
the structure of the landscape and the dispersal capabilities of the focal
species \citep{Gustafson.1996}. For species that disperse short distances
through homogeneous environments, few simulations may suffice to gauge
connectivity, whereas for species that disperse over long distances through
heterogeneous habitats, a large number of simulations will be required to
sufficiently explore the spectrum of possible routes. Finally, it may often
suffice to extract covariates at each step's start or endpoints, thus
considerably speeding up simulation times \citep{Signer.2017}.

### 2023-03-14 Chapter_2/04_Manuscript/Manuscript.tex
various updates

\subsection{Spatial Habitat Layers}
We attempted to represent each spatial covariate at the highest temporal
resolution available, which resulted in various degrees of temporal accuracy. In
order to generate spatial layers that depict the flood extent at different
points in time, we ... \Cref{FloodExtent}.

>>>>

\subsection{Spatial Habitat Layers}
In order to generate spatial layers that depict the flood extent at different
points in time, we ... \Cref{FloodExtent}.
